---
title: "Red Wine Quality"
author: Kyle Hansen
date: February 17, 2021
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: readable
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(dplyr)
library(gridExtra)

```

```{r echo=FALSE, Load_the_Data}
# Load the Data
rw <- read.csv('wineQualityReds.csv')

rw$quality <- factor(rw$quality, ordered = TRUE)
```

This report explores a dataset containing red wines and variables on their 
chemical properties.  Also included are 3 wine experts ratings for each wine on
a scale of 0 (very bad) to 10 (very excellent)

# Univariate Plots Section


```{r echo=FALSE, Univariate_Plots}
str(rw)
```


```{r echo = FALSE}
summary(rw)
```

Here is the list of variables:
```{r echo= FALSE}
names(rw)
```


The dataset contains 13 variables, with 1,599 observations.

```{r echo= FALSE, functions}

#Helpful Functions to speed up the univariate plotting

func_summary <- function(variable){
  print(summary(rw[[variable]]))
}

func_histogram <- function(variable, ...){
  ggplot(data = rw, aes_q(as.name(variable)))+
    geom_histogram(...)
}

summary_and_histogram <- function(variable, ...){
  func_summary(variable)
  func_histogram(variable, ...)
}


```

### Fixed Acidity
```{r echo = FALSE, fixed.acidity}
summary_and_histogram('fixed.acidity', binwidth= .25, fill = 'blue') +
  scale_x_continuous(breaks= 4:16)
```

The majority of wines have an acidity between 7.10 and 9.20 $g/dm^3$. The median 
fixed acidity of the red wines in the dataset is 7.90 $g/dm^3$. The distribution 
of fixed acidity is right skewed. There are outliers in the higher range ~ >15.


### Volatile Acidity
```{r echo = FALSE}
summary_and_histogram('volatile.acidity', binwidth= .025, fill = 'blue') +
  scale_x_continuous(breaks= seq(0, 1.7, .1))
```

### Citric Acid
```{r echo = FALSE}
summary_and_histogram('citric.acid', binwidth= .025, fill = 'blue') +
  scale_x_continuous(breaks= seq(0, 1, .125))
```

Most of the wines have 0 $g/dm^3$ of citric acid and the amount of citric acid 
is always found in very small quantities. The distribution is right skewed with
a couple of smaller peaks. The peaks are at 0.250 and 0.500 which may hint at 
some bimodal behavior. A single red wine appears to be an outlier on the right 
side with 1 $g/dm^3$ of citric acid.

### Residual Sugar
```{r echo = FALSE}
summary_and_histogram('residual.sugar', binwidth= .25, color = 'black', fill = 'blue') +
  scale_x_continuous(breaks= 0:16) +
  coord_cartesian(xlim = c(1, 9))

```

The distribution of has a median of 2.2. The distribution is skewed right with a 
long tail to the right. There are many very small bars on the right side of the
peak, while the majority amount of residual sugar in wine falls between 1.5 - 
2.5 $g/dm^3$.

### Chlorides

```{r echo= FALSE, chlorides}
summary_and_histogram('chlorides', binwidth= .01, color = 'black', fill = 'blue') +
  scale_x_continuous(breaks= seq(0, .6, .1))

```

The median value of chlorides in the red wine is 0.079 $g/dm^3$. The 
distribution looks normal near its  peak but has a long right tail with small 
counts of wines with values going until 0.611 $g/dm^3$.

### Free Sulfur Dioxide

```{r echo= FALSE, free.sulfur.dioxide}
summary_and_histogram('free.sulfur.dioxide', binwidth= 1, color = 'black' , fill = 'blue') +
  scale_x_continuous(breaks= seq(0, 75, 5))

```

The distribution of free sulfur dioxide is right skewed and has a median value 
of 14 $mg/dm^3$. The right tail extends until 72 with a gap between 57 and 66.

### Total Sulfur Dioxide

```{r echo = FALSE, total.sulfur.dioxide}
summary_and_histogram('total.sulfur.dioxide', binwidth= 5, color= 'black', fill = 'blue') +
  scale_x_continuous(breaks = seq(0, 300, 25))

```

This distribution is right skewed with a median value of 38 $mg/dm^3$. There is
a large gap between 165 and 278 in the right tail with only 2 wines with a 
greater concentration equal to 278 or more.

### Denisty

```{r echo = FALSE, density}
summary_and_histogram('density', binwidth= .00025, color = 'black', fill = 'blue') +
  scale_x_continuous(breaks= seq(0.99, 1.005, 0.0025))

```

The density of wines varies little, with most values falling between 0.9956 and
0.9967. The distribution is almost balanced and has median value of 0.9968 
$g/cm^3$. The density is close to the density of water.


### pH
```{r echo= FALSE, pH}
summary_and_histogram('pH', binwidth= 0.025, color='black', fill = 'blue') +
  scale_x_continuous(breaks= seq(0, 4.5, 0.1))

```

The wines in the dataset have low pH. The distribution looks symmetrical, but 
could also be considered bimodal with both peaks very close to each other.  The
median pH value is 3.31 and most wines have a pH between 3.21 and 3.4.

### Sulphates

```{r echo= FALSE, sulphates}
summary_and_histogram('sulphates', binwidth= .02, color = 'Black', fill = 'blue') +
  scale_x_continuous(breaks= seq(0, 2, .2))

```

The distribution is slightly skewed right. Outliers in the right tail have about 
2 $g/dm^3$ of sulphates and the median value of sulphates is 0.62. Most wines 
have a sulphate concentration between 0.55 and 0.65.


### Alcohol
```{r echo = FALSE, alcohol}
summary_and_histogram('alcohol', binwidth= .1, color = 'black', fill = 'blue') +
  scale_x_continuous(breaks= seq(8, 15, 1))

```

The distribution is right skewed and there seems to be a natural border on the
left side at about 9%, most likely due to wine having a minimum amount of 
alcohol needed in it in order for a wine to be considered a wine. The high peak 
of the distribution is at 9.5 % alcohol and the median value is 10.20%, while 
the maximum amount of alcohol present in the dataset is 14.90%.

### Quality
```{r echo = FALSE, wine_Quality}

ggplot(aes(x= quality), data= rw) +
  geom_bar(color = 'black', fill = 'blue')

```

From the graph above, the distribution of the quality ratings of the wine appear
to be normal with the majority of wines rated at average quality (between 5-6). 
No wines got a rating less than 3 or greater than 8.




# Univariate Analysis

### What is the structure of your dataset?
The structure of the dataset entails 12 variables and 1599 observations.  11 
variables are different physicochemical test results and 1 variable corresponds 
to the quality rating from an expert wine panel.  Each observation signifies a 
red wine sample.

### What is/are the main feature(s) of interest in your dataset?
The main feature of interest in the data set is the quality rating.

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?
I think most of the physicochemical test results may help support the 
investigation into the quality rating given by the expert panels, since all of
them are different characteristics to the wine that could have an affect on the 
flavor.  For exmaple, density depends on the percent of alcohol and sugar 
content, and pH results show how acidic the wine is.  Both can impact the 
flavor of the wine.

### Did you create any new variables from existing variables in the dataset?
No new variables created.


### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?
No unusal distributions or missing values seen in the dataset.  The dataset 
given was already tidy and clean, and had no need for any adjustments.


# Bivariate Plots Section


```{r echo= FALSE}
boxplot_compare <- function(variable){
  ggplot(data= rw, aes_q(x = ~quality, y= as.name(variable))) +
    geom_boxplot() +
    geom_jitter() +
    geom_line(aes(group = 1), stat = "summary", fun = median, color = "red", size = 1, alpha = .75)+
    stat_summary(fun= mean, geom= 'point', shape= 7, color = 'blue')
  
}

comp_mean <- function(variable){
  print(paste("Mean of", variable, "by quality:"))
  print(by(rw[[variable]], rw$quality, mean))
}

boxplot_and_mean <- function(variable){
  comp_mean(variable)
  boxplot_compare(variable)
}
```


## Alcohol vs. Quality Rating

```{r echo = FALSE}

boxplot_and_mean('alcohol')


```

Except for the downward slope from quality rating 4 to 5, the higher the alcohol 
content in the wine, the higher the rating the wine was given.


## Chlorides vs Quality Rating

```{r echo = FALSE}

boxplot_and_mean('chlorides')

```

The small relation shows that the less chlorides, the higher the quality rating.

## Sulphates vs Quality Rating
```{r echo = FALSE}
boxplot_and_mean('sulphates')
```

The graph shows the higher the sulphates concentration, the higher the quality
rating of the wine.

## Citric Acid vs Quality Rating
```{r echo = FALSE}
boxplot_and_mean('citric.acid')
```

The graph shows the higher the citric acid in the wine, the higher the quality
rating of the wine.  


## Volatile Acidity vs Quality Rating

```{r echo = FALSE}

boxplot_and_mean('volatile.acidity')

```

The lower the voatile acidity in wines seem to have an affect on having a higher
quality rating

## pH vs Quality Rating
```{r echo = FALSE}
boxplot_and_mean('pH')

```

There seems to be a higher quality rating with lower pH levels, which translates 
to the more acidity in alcohol the higher the quality.  Checking correlations of
acidity and pH should help verify this.

## Ph and Acidity Correlations

```{r echo= FALSE, warning= FALSE}


p1 = ggplot(aes(x= pH, y= fixed.acidity), data= rw) +
  geom_point(alpha = .25) +
  scale_y_log10()

p2 = ggplot(aes(x= pH, y= citric.acid), data= rw) +
  geom_point(alpha = .25) +
  scale_y_log10()

p3 = ggplot(aes(x= pH, y= volatile.acidity), data= rw) +
  geom_point(alpha = .25) +
  scale_y_log10()


grid.arrange(p1, p2, p3)



```

According to the above graphs, fixed acidity accounts for most of the acid found 
in the wines, and has the most correlation of the pH levels increasing, with the
lower amount of acid.  A slight, similar relation is seen in citric acids, but 
since there is so little citric acid in the wines, the relationship isnt as 
strong.  And volatile acidity seems to have just only a very slight positivie
correlation if any.


# Bivariate Analysis


### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in dataset?

The higher quality rating has a stronger relationship with alcohol content, 
sulphates, citric acid, and volatile acidity.  Here is the correlation
to quality for the variables:

```{r echo= FALSE}
#Creating a funciton to input the variables to get the correlation to quality
cor_test <- function(x, y) {
  return(cor.test(x, as.numeric(y))$estimate)
}

quality_cor <- c(
  cor_test(rw$alcohol, rw$quality),
  cor_test(rw$sulphates, rw$quality),
  cor_test(rw$citric.acid, rw$quality),
  cor_test(rw$volatile.acidity, rw$quality)
)

names(quality_cor) <- c('Alcohol/Quality', 'Sulphates', 'Citric Acid', 'Volatile Acidity')

quality_cor


```

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

I observed the relationship between the pH levels with the acidity levels.  I 
wasnt suprised to find that fixed acidity had the most affect on pH levels due 
to there is more of it in the wines compared to the citric or volatile acids.


### What was the strongest relationship you found?
The strongest relationship with the quality rating is the alcohol content, based 
on the correlation coefficients.

# Multivariate Plots Section



## Volatile Acidity, Alcohol, and Quality Rating

```{r echo=FALSE, Multivariate_Plots}
#Function for Multivatiarte Analysis to allow input of 2 variable 
#properties of wine to compare with the Quality Rating of that wine

scatter_with_quality <- function(x, y){
  ggplot(rw, aes_q(x = as.name(x), y = as.name(y), color = ~ quality))+
    geom_point(alpha = 0.75, size = 2)+
    scale_color_brewer(type = "div", palette = "Spectral")
}

scatter_with_quality("alcohol", "volatile.acidity")


```

The data shows, the the lower the volatile acidity and higher the alcohol 
content, the higher the quality rating.  The middle quality rating wines is more
spread out.

## Citric acid, alcohol, and quality rating

```{r echo=FALSE}

scatter_with_quality("alcohol", "citric.acid")

```

The data shows, that with an increase in citric acid and alcohol, the quality 
rating tends to be higher.  The quality ratings of 5 and 6 are showing to have a 
wide range of citric acid levels with lower alcohol.

## Alcohol, Sulfphates, and Quality Rating

```{r echo=FALSE}

scatter_with_quality("alcohol", "sulphates")

```

The data shows that when sulphates are about .75 and up and alcohol content is
about 11% and greater, the quality rating is higher.



# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

The relationsihps observered were between the greatest correlating attributes
with the quality rating giving by the experts.  The data shows that the wines 
with higher alcohol and lower acidity were generally given higher quality
ratings.  Also, along with higher alcohol content, the higher amounts of citric
acid in a wine yielded higher quality ratings.  Also, as the sulphates amount 
increased with the alcohol content increasing, the quality rating was higher.


------

# Final Plots and Summary


### Plot One
```{r echo=FALSE, Plot_One, fig.width=10, fig.height=6}

boxplot_compare("volatile.acidity")+
  xlab("Expert Quality Rating")+
  ylab("Volatile Acidity (g / dm^3^)")+
  ggtitle("Higher acetic acid content = lower quality rating")+
  theme(text = element_text(size = 20))

```

### Description One
With this plot, one can see the distribution of the acetic acid amount in the 
wine across quality ratings.  The boxplot shows the quantile boundaries and  
mean values (blue dots), while the overlapping dots shows the distribution of 
wines in the quality ratings. There is an unbalanced amount between the middle 
ratings and the extreme ratings. There are many more middle quality rated wines 
than there are low and high quality rated. The red line follows and connects 
the median values and shows visually the decreasing trend of volatile acidity 
amount with quality. Wines with lower volatile acidity have a higher rating in 
quality.

### Plot Two

```{r echo=FALSE, Plot_Three, fig.width=10, fig.height=6}
scatter_with_quality(x = "alcohol", y = "volatile.acidity")+
  xlab("Alcohol (% by Volume)")+
  ylab("Volatile Acidity (g / dm^3^)")+
  ggtitle("Quality Rating is affected by volatile acidity and alcohol")+
  labs(color = "Quality Rating")+
  theme(text = element_text(size = 20))


```


### Description Two
Is has been seen that alcohol and volatile acidity had a large influence on the 
quality rating of the wines. In this plot, we see the combined effects of these
two variables on the quality rating. The red wines with high volatile acidity 
and low alcohol have a lower quality rating. Also at low volatile acidity and 
low alcohol, the wines tend to have a middle of the road quality rating. And at
low volatile acidity and high alcohol content, there are higher quality rated 
wines.


### Plot Three
```{r echo=FALSE, Plot_Two, fig.width=10, fig.height=6}

boxplot_compare("alcohol")+
  xlab("Expert Quality Rating")+
  ylab("Alcohol (% by Volume)")+
  ggtitle("Higher alcohol content = higher quality")+
  theme(text = element_text(size = 20))

```

### Description Three

There is a  big impact on quality of wines due to aclohol content. For the 
quality ratings of 3 to 5, the impact isnt very high and this could be due to
another variable. But from the quality ratings from 5 to 8, there is a sharp 
increase in the alcohol content. The general trend is that wines with greater
alcohol content have a greater quality rating.

------

# Reflection
I thought this was a fun, and interesting project to put all the knowledge 
learned of R plotting functionality and exploring a real dataset. The dataset
was already very well organized and cleaned without any missing data. The only
downside of the dataset was that there were many more wines at the middle 
quality levels than at the low and high ends.

The first challenge was to choose which direction to start exploring the data. 
But luckily, the dataset description file already hints at some variables to 
look at. As an example in the description file, it explains that citric acid can 
add "freshness and flavor" to wines, while acetic acid can add an unpleasant 
vinegar taste. This shows that its important to have specific detailed knowledge
while performing a data analysis. Without that understanding, one can spend too
much time exploring the data in an unhelpul and prooductive way. 


I think this analysis could be more enriched in the future if we could bring in
different wine experts and give their reviews on the same wines, and see if they
correlate and line up with the first experts ratings.  I think it would be 
interesting to see if there is a strong correlation to a bias pallete when
deciding the wine qualities and if they allign with the original ratings of each
wine.  

Finally, since the main trends have been identified in the data, prediction 
models could be built to see how the trends can be used to predict red wine 
quality based on the physicochemical attributes.

# Reference
P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
  Modeling wine preferences by data mining from physicochemical properties.
  In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.


https://cran.r-project.org/web/packages/egg/vignettes/Ecosystem.html